Approximating the matrix exponential derivative of the CTMC generator matrix, with application to global SARS-CoV-2

### 2023-03-14 manuscript/main.tex
another paragraph

@@

Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate Metropolis-type \citep{metropolis1953equation} proposals far away from the current Markov chain state.  On the one hand, this approach to proposal generation helps reduce autocorrelation between chain states and is particularly helpful within higher-dimensional state spaces.  On the other hand, numerical integration of Hamilton's equations requires repeated evaluation of the Hamiltonian potential energy's gradient, and these repeated floating-point operations may become computationally burdensome.   Using HMC for Bayesian inference, the potential energy takes the form $u(\ttheta)=-\log \pi(\ttheta)$, where $\pi(\cdot)$ is the posterior density function $\pi(\ttheta):=p(\ttheta|\Y)$ of parameter $\ttheta\in\mathbb{R}^D$ given observed data $\Y$.  In this context, a single iteration of the the St\"{o}rmer-Verlet (velocity Verlet) or leapfrog integrator \citep{leimkuhler2004simulating}, beginning at time $\tau$ and with $\epsilon>0$ small, takes the form
\begin{align}\label{eq:leap}
	\mom\left(\tau + \frac{\epsilon}{2}\right) &= \mom(\tau) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau))  \\  \nonumber
	\ttheta(\tau+\epsilon) &= \ttheta(\tau) + \epsilon \, \MM^{-1}\mom(\tau+\frac{\epsilon}{2}) \\ \nonumber
	\mom(\tau + \epsilon) &= \mom\left(\tau+\frac{\epsilon}{2}\right) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau+\epsilon))  \, ,
\end{align}
where $\mom \sim \mbox{N}_D(\Zero,\MM)$ is the Hamiltonian system's Gaussian-distributed momentum variable.  
Different strategies aim to speed up the leapfrog integrator's many log-posterior gradient evaluations $\nabla \log \pi (\ttheta)$ in different ways, as these numerical routines often represent the algorithm's computational bottleneck.  In model-specific contexts, \citet{holbrook2021massive,holbrook2022bayesian,holbrook2022viral} develop parallelization strategies, and \citet{ji2020gradients} develop dynamic programming techniques, to accelerate $\nabla \log \pi (\ttheta)$.

A small body of literature considers another approach by recognizing

>>>>

 We are interested in efficiently approximating the directional derivative of the matrix exponential
\begin{align*}
	e^{t\QQ} = \sum_{k=0}^\infty \frac{t^k\QQ^k}{k!}
\end{align*}
for arbitrary $t>0$ using the formula \citep{najfeld1995derivatives}
\begin{align}\label{eq:deriv}
	\nabla_{\JJ} e^{t\QQ}  =  e^{t\QQ}  \sum_{k=0}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \, .
\end{align}
Here, $\JJ$ gives the direction and $\{\JJ,\QQ^k\}$ is the matrix commutator series satisfying the recursion
\begin{align*}
	\{\JJ,\QQ^0\} &= \JJ \, , \\ 
	\{\JJ,\QQ^k\} &= [\{\JJ,\QQ^{k-1}\} , \QQ]  = \{\JJ,\QQ^{k-1}\} \QQ - \QQ \{\JJ,\QQ^{k-1}\} \, .
\end{align*}
Thus, the first-order approximation to derivative \eqref{eq:deriv} is 
\begin{align}\label{eq:firstOrder}
	\widetilde{\nabla}_{\JJ} e^{t\QQ}= t e^{t\QQ}\JJ \, .
\end{align}

@@

Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid,neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate Metropolis-type \citep{metropolis1953equation} proposals far away from the current Markov chain state.  On the one hand, this approach to proposal generation helps reduce autocorrelation between chain states and is particularly helpful within higher-dimensional state spaces.  On the other hand, numerical integration of Hamilton's equations requires repeated evaluation of the Hamiltonian potential energy's gradient, and these repeated floating-point operations may become computationally burdensome.   Using HMC for Bayesian inference, the potential energy takes the form $u(\ttheta)=-\log \pi(\ttheta)$, where $\pi(\cdot)$ is the posterior density function $\pi(\ttheta):=p(\ttheta|\Y)$ of parameter $\ttheta\in\mathbb{R}^D$ given observed data $\Y$.  In this context, a single iteration of the the St\"{o}rmer-Verlet (velocity Verlet) or leapfrog integrator \citep{leimkuhler2004simulating}, beginning at time $\tau$ and with $\epsilon>0$ small, takes the form
\begin{align}\label{eq:leap}
	\mom\left(\tau + \frac{\epsilon}{2}\right) &= \mom(\tau) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau))  \\  \nonumber
	\ttheta(\tau+\epsilon) &= \ttheta(\tau) + \epsilon \, \MM^{-1}\mom(\tau+\frac{\epsilon}{2}) \\ \nonumber
	\mom(\tau + \epsilon) &= \mom\left(\tau+\frac{\epsilon}{2}\right) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau+\epsilon))  \, ,
\end{align}
where $\mom \sim \mbox{N}_D(\Zero,\MM)$ is the Hamiltonian system's Gaussian-distributed momentum variable.  
Different strategies aim to speed up the leapfrog integrator's many log-posterior gradient evaluations $\nabla \log \pi (\ttheta)$ in different ways, as these numerical routines often represent the algorithm's computational bottleneck.  In model-specific contexts, \citet{holbrook2021massive,holbrook2022bayesian,holbrook2022viral} develop parallelization strategies, and \citet{ji2020gradients} develop dynamic programming techniques, to accelerate $\nabla \log \pi (\ttheta)$.

A small body of work considers another approach by replacing $\nabla \log \pi (\ttheta)$ with a suitable approximation $\widehat{\nabla} \log \pi (\ttheta)$ and recognizing that the modified \eqref{eq:leap} continues to satisfy path reversibility and volume preservation, two essential ingredients for well-specified HMC. It follows immediately that the resulting `surrogate trajectory HMC' continues to sample the correct target distribution $\pi(\cdot)$.  Nonetheless, the acceptance rates and overall efficiency of such samplers may suffer when approximations are poor.   The majority of surrogate HMC methods first obtain a small sample of exact gradient evaluations and then use some model to interpolate: \citet{zhang2017precomputing} assume the approximate gradient follows a piecewise constant form across a grid; \citet{rasmussen2003gaussian,lan2016emulation} construct approximations using Gaussian processes; and \citet{zhang2017hamiltonian,li2019neural} do the same using neural networks.  But an even simpler approach to surrogate HMC may be appropriate when gradients have series representations.   \textcolor{red}{MAGEE} consider phylogenetic models that involve CTMC priors and show that the first-order approximation \eqref{eq:firstOrder} of the matrix exponential derivative \eqref{eq:deriv} performs remarkably well within surrogate HMC, achieving an over 30-fold efficiency gain compared to random-walk Metropolis for a 14-state model with over 180 model parameters.  Here, we demonstrate similar strong performance of the first-order approximation within surrogate HMC for a 44-state model with almost 1900 model parameters.
